\documentclass[10pt,a4paper,oneside]{letter}

\usepackage[includeheadfoot,margin=2cm,head=28pt,foot=20pt]{geometry}
 % \usepackage{fullpage}
 \setlength{\parindent}{0pt}

\newcommand{\entrylabel}[1]{\mbox{#1}\hfil}

\newenvironment{myitemize}
  {\begin{list}{}
    {\renewcommand{\makelabel}{\entrylabel}
    \setlength{\labelwidth}{1ex}
    \setlength{\leftmargin}{0ex}
    \setlength{\parsep}{1ex}
  }}
  {\end{list}}



\address{
{Professor Vladimir Voevodsky} \\
{School of Mathematics} \\
{Institute for Advanced Study} \\
{Princeton, USA}}

\date{July 7th, 2014}

\signature{Vladimir Voevodsky}


\begin{document}



\begin{letter}{
EPSRC \\
Polaris House, \\
North Star Avenue, \\
Swindon SN2 1ET\\
United Kingdom}

\opening{TO WHOM THIS MAY CONCERN}

I am writing to express my desire to collaborate on the research proposal
``Homotopy Type Theory: Programming and Verification'', submitted to
EPSRC by Thorsten Altenkirch, Nicola Gambino, Neil Ghani and Conor
McBride. 

%Who am I
I am Full Professor at the School of Mathematics of the Institute for
Advanced Study in Princeton (USA).  Until 2010, my research has
focused primarily on algebraic geometry. I made several fundamental
contributions to the subject, for which I was awarded of a Fields
Medal in 2002. From 2006 onwards, my research has shifted
progressively towards the foundations of mathematics and theoretical
computer science. My discovery of a homotopy-theoretic model of type
theory led me to introduce the Univalence Axiom and to formulate the
Univalent Foundations research programme, which seeks to develop a new
approach to the foundations of mathematics and facilitate the
computer-assisted verification of mathematical proofs. I have been
leading the development of this programme since its inception; in
particular, I was one of the coordinators of the special year on this
topic at the Institute for Advanced Study in 2012/13 and I am
currently one of the leading investigators of the recently-awarded
$\$7.5$M grant on Homotopy Type Theory from the the US Air Force
Office for Scientific Research.

%The proposal is excellent as the research must be done, it is distinctive
I am very positive about the proposal, especially because it brings
the ideas of Univalent Foundations to bear upon the development of
programming languages and programming verification tools, thus
complementing ideally the current research on Univalent Foundations,
which focuses mainly on the connections with homotopy theory and the
applications to formalization of mathematics.  It is thus distinctive
in its goals. Furthermore, the proposal builds convincingly on
previous experience and results of the applicants. In particular, the
work of Dr Altenkirch and Dr McBride on Observational Type Theory
contains already important ideas on the treatment of equality in type
theory, while the work of Dr Gambino on homotopy inductive types shows
the usefulness of the connections between type theory and homotopy
theory. In particular, I have been in close contact with Dr Gambino
and Dr Altenkirch, who have spent extended periods visiting me at the
Institute for Advanced Study.  The relationship between logical
relations and Homotopy Type Theory definitely needs greater
understanding so Prof Ghani's presence is ideal. Finally, Dr McBride's
experience as an implementor of dependently typed programming
languages rounds off a very well-balanced team which possesses the
experience, diversity and ability needed to develop successfully the
proposed research. The areas of expertise of the team - and the ideas
being deployed - means the proposal is also distinctive in terms of
the techniques used to achieve the overall goals. When taken together
with the collaboration of the leaders of the field, this makes the
chances of success very strong.





%My work package
 I would very happy to collaborate with with the
research team.  Our collaboration will focus on the
research goals of Work Package 2. {\bf MORE TO BE ADDED}.
Subject to approval by the Institute for Advanced Study, I would be happy
to host 4 one-week visits by members by the research team, as well as
keep in regular contact by the usual academic means.





\closing{With best regards,}

\end{letter}

\end{document}


