\documentclass[a4paper,11pt]{article}

\usepackage[top=2cm, bottom=2cm, left=2cm, right=2cm]{geometry}

\usepackage{mathptmx}
\usepackage{epsf}           %\input{epsf}
\usepackage{amsfonts}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{url}
%\usepackage[dvips]{graphics}
\usepackage[dvips,pdftex]{graphicx}
\usepackage[dvips,all]{xy}
\usepackage{multicol}
\usepackage{natbib}
\setlength{\bibsep}{0.0pt}

\usepackage{microtype}
\usepackage{hyperref}

%\newlength{\extraplusheight}
%\newlength{\extrapluswidth}
%\setlength{\extraplusheight}{4.7cm}
%\setlength{\extrapluswidth}{4.7cm}
%\addtolength{\textwidth}{\extrapluswidth}
%\addtolength{\textheight}{\extraplusheight}
%\addtolength{\oddsidemargin}{-.5\extrapluswidth}
%\addtolength{\evensidemargin}{-.5\extrapluswidth}
%\addtolength{\topmargin}{-0.5\extraplusheight}
\setlength{\parindent}{0 pt}
\setlength{\parskip}{.5ex}
%\documentclass[a4paper]{article}

\usepackage{pgfgantt}
  

\begin{document}
%\hspace{-3cm}
\thispagestyle{empty}

{
\centering
{\Large \bf HoTT : Programming and Verification Workplan}
\medskip

\setganttlinklabel{s-s}{}
\setganttlinklabel{f-f}{}
\setganttlinklabel{s-f}{}

\begin{tikzpicture}[y=0.5cm]  
\begin{ganttchart}[x unit=1.75cm,y unit chart=.5cm,y unit title=.5cm,vgrid,
canvas/.append style={fill=none, draw=black!5, line width=.75pt},
hgrid style/.style={draw=black!5, line width=.75pt},
vgrid={*1{draw=black!4,dashed,line width=.5pt},*1{draw=black!5, line width=.75pt}},
  title/.style={draw=none, fill=black!5},
%  title label node/.append style={below=0pt},
  include title in canvas=false,
  bar label font=\mdseries\small\color{black!70},
  bar label node/.append style={left=.5cm},
  bar/.append style={draw=none, fill=magenta!63},
%  bar incomplete/.append style={fill=barblue},
%  bar progress label font=\mdseries\footnotesize\color{black!70},
%  group incomplete/.append style={fill=groupblue},
%  group left shift=0,
%  group right shift=0,
%  group height=.5,
%  group peaks tip position=0,
%  group label node/.append style={left=.6cm},
%  group progress label font=\bfseries\small,
 link/.style={-latex, line width=0.5pt, black},
 link label font=\scriptsize\bfseries,
 link label node/.append style={below left=-2pt and 0pt},
 title height=.75, title top shift=0,
% title label anchor={below=-1.5ex},
%,
% bar top shift=-0.1, 
 bar height=0.5
  ]{1}{8}
  \gantttitle{2015}{2}
  \gantttitle{2016}{2}
  \gantttitle{2017}{2}
  \gantttitle{2018}{2}\\
  \ganttbar{WP1}{1}{3}\\
  \ganttbar{WP2}{2}{5}\\
  \ganttbar{WP3}{4}{6}\\
  \ganttbar{WP4}{6}{8}\\
  \ganttbar{WP5}{1}{4}\ganttbar{WP5}{7}{8}\\
  \ganttbar{WP6}{3}{6}\\
  \ganttbar{WP7}{5}{8}\\
  \ganttbar{WP8}{7}{8}
  % \ganttlink[link type=s-s]{elem0}{elem1}
  % \ganttlink[link type=s-s]{elem0}{elem2}
  % \ganttlink[link type=s-s]{elem0}{elem4}
  % \ganttlink[link type=s-s]{elem1}{elem2}
  % \ganttlink[link type=s-s]{elem1}{elem4}
  % \ganttlink[link type=s-s]{elem1}{elem5}
  % \ganttlink[link type=s-s]{elem2}{elem3}
  % \ganttlink[link type=s-s]{elem5}{elem4}
\end{ganttchart}
\end{tikzpicture}

}

\bigskip

{\bf Relationships between work packages.} The following 
relationships between work packages feed into the design of the above
Gantt chart. WP1,5 start immediately as both have no prerequisites. We
start WP2 early as it is a key work package and we already have
ideas to pursue within Altenkirch's Cubical Type Theory. An early
start to WP2 also minimises the risk that unforeseen difficulties
within this key work package holds up progress elsewhere. As WP1
progresses, its results will feed into WP2. Similarly, another key
work package is our core language of WP6 and so we minimise risk by
starting it as soon as possible --- since WP6 is built on WP2, we
start WP6 a year into the project. WP3 needs to be consistent with our
syntactic and semantic foundations and so starts after WP1,2 have had
time to deliver results. As results arise in WP3, they will be fed
into WP6. As WP7 builds upon WP6 and WP3, it starts after those work
packages have had time to deliver results. WP4 builds upon WP3 as is
reflected in its start date. Once most of our tools are
developed, we start WP8.

{\bf Duration.} The longest work package is WP5 as this is where our
formal verification takes place, and hence is our broadest in terms of
scope. We want to formally verify as many results as possible about
our foundations, programming language tools, and impact generating
case studies.  As a result this is the longest work package which
(with a break for language development) will be undertaken at both the
beginning and the end of the project. WP2,6,7, while less broad, are
equally fundamental to our overall goals and so they are allocated 2
years each. The other work packages are more focussed and
so are given 1.5 years apart from WP8 which we think will only need 1
year as WP8 is the most tightly scoped work package.

Nevertheless, there is flexibility in these lengths to either extend
or shorten work packages as the pace of progress within them becomes
evident. For example, if our work on WP6,7 delivers the required
platform, we will start work on WP8 earlier or formally verify more in
WP5.  Our planning of the timing and duration of these work packages
has ensured that both RAs will not have to work on more than two work
packages at a time --- the one exception is in Yr4 but
there we expect the Nottingham RA to be helping in WP5 by formalising
results of WP4. The above durations have also been
devised to reduce overall risk, in that the largest risk in the
overall proposal is the amount of engineering work required by
programming language and verification tools. This risk is mitigated by
McBride's presence in the team and by separating the
more more modest goal of a core system (WP6) from the more ambitious
goal of a high level system (WP7) and thereby allocating 50$\%$ of the
time of the Strathclyde RA to WP6 and WP7. Further, we mitigate risk 
by aiming in both WP6 and WP7 at a fundamental innovation and not a
fully-fledged system able to compete with Coq or Agda, as that would
be unrealistic.

{\bf Leadership.} WP1 will be jointly led by Nottingham and Leeds as
both Gambino and Altenkirch have considerable experience there. Having
joint leadership will not be a problem due to the proximity of these
sites and because Gambino and Altenkirch successfully collaborated on
the HoTT book during the Special Year on HoTT in Princeton. Indeed,
this will be advantageous as the Nottingham RA will get to work
closely with Gambino. Ghani's experience in category theory will make
him a secondary contributor on WP1. WP2--4 will be led at Nottingham
as Altenkirch has significant experience in all of these areas
including a track record of collaborating with Microsoft. All
investigators have experience in type theory and data types, and hence
are secondary contributors to WP2,3. Ghani's experience in logical
relations will also be deployed here. Ghani and McBride will similarly
contribute to WP4 because of their work on effects and with
Microsoft. WP5--8 will be led at Strathclyde and McBride will take the
lead on WP5-7 as he has the most expertise in these areas, while Ghani
will take the lead on WP8 because of his experience working with
Kennedy on Units of Measure. Altenkirch will be a secondary
contributor to WP5--7 given his experience in programming language
implementation and the formalisation of type theory and Gambino will
be a secondary contributor to WP5 as he has experience there.

\end{document}
