\documentclass[a4paper,11pt]{article}

\usepackage[top=2cm, bottom=2cm, left=2cm, right=2cm]{geometry}

\usepackage{mathptmx}
\usepackage{epsf}           %\input{epsf}
\usepackage{amsfonts}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{url}
 \usepackage[dvips]{graphics}
\usepackage[dvips,all]{xy}
\usepackage{multicol}

%\newlength{\extraplusheight}
%\newlength{\extrapluswidth}
%\setlength{\extraplusheight}{4.7cm}
%\setlength{\extrapluswidth}{4.7cm}
%\addtolength{\textwidth}{\extrapluswidth}
%\addtolength{\textheight}{\extraplusheight}
%\addtolength{\oddsidemargin}{-.5\extrapluswidth}
%\addtolength{\evensidemargin}{-.5\extrapluswidth}
%\addtolength{\topmargin}{-0.5\extraplusheight}
\setlength{\parindent}{0 pt}
\setlength{\parskip}{1ex}

\newcommand{\Int}[1]{[\![ #1 ]\!]}
\newcommand{\malign}[1]{\begin{array}[t]{@{}l@{\;}l@{}l@{}} #1 \end{array}}
\newcommand{\logrel}[2]{\Delta_{#1,#2}}
\newsavebox{\fminibox}
\newenvironment{fminipage}
 {\begin{lrbox}{\fminibox}\begin{minipage}{8cm}\vspace*{-2ex}}
 {\\[-2ex]\vspace*{-2ex}\end{minipage}\end{lrbox}\noindent\centerline{\fbox{\usebox{\fminibox}}}\vspace{0.5ex}}   

\setlength{\parindent}{0.15in}
%\setlength{\parskip}{0.3ex}

% Discourage unnecessary hyphenation.
\sloppy\hyphenpenalty 4000

\newcommand{\ra}{\rightarrow}
\newcommand{\A}{\mathcal{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\Set}{\mbox{{\sf Set}}}
\newcommand{\Nat}{\mathit{Nat}}
\newcommand{\Alge}[1]{\mathit{Alg}_{#1}}
\newcommand{\hash}{\#}



\begin{document}

\thispagestyle{plain}
\begin{center}
  {\Large \bf Homotopy Type Theory: Programming and Verification\\
\vspace{0.2in}
  Justification of Resources}\\[1ex]

\vspace*{-0.1in}

\rule{160mm}{.5mm}\\[2ex]
\end{center}

\noindent While EPSRC is principally interested in the quality of research,
costs must be justified. It is argued below that i) all four investigators
are required for the proposed research because they have
distinct-but-necessary skills; ii) two RAs are required because of the
breadth and volume of the work proposed; iii) the duration of the
grant is appropriate given the volume of the work involved and the
fast moving nature of HoTT research; iv) the
hours per week spent on the grant by the investigators are required;
and 
v) the travel budget is needed to underpin the planned collaborations,
workshops, and attendance at conferences.

\vspace{0.02in}

\noindent {\bf The Investigators:} All four investigators have
excellent track records in the areas of the proposed research: i)
Altenkirch is an expert in type theory and the syntactic presentation
of HoTT --- in particular his work on OTT, Cubical Type Theory,
Containers, modelling type theory in type theory and the experimental
programming language $\Pi\Sigma$ has direct relevance to WPs
2,3,5,6,7; ii) Gambino is an expert in category theory and the
semantic foundations of HoTT with his work on {\bf ??}  and dependent
polynomials having direct relevance to WPs 1,3; iii) Ghani is an
expert on the categorical foundations of programming languages with
his work on logical relations, containers, effects and units of
measure being particularly relevant to WP1, 2, 3, 4 and 8; and iv)
McBride is an expert on the definition and implementation of
programming languages with his work on containers, OTT, effects and
Epigram being particularly relevant to WPs 3,4,6,7. These skills
directly map onto the management of the proposed research as
follows. WP1 will be lead by Leeds as Gambino is the strongest
researcher on the semantics of HoTT, while WPs 2,3 and 5 will be lead
by Nottingham as they reflect strengths of Dr Altenkirch on the type
theoretic foundations of programming languages. Finally WPs 4,6,7 and
8 will be lead by Strathclyde given their expertise in dependently
typed programming. Ghani's role - apart from specific scientific
contributions outlined above --- will be to knit the theoretical
research done at Nottingham and Leeds with the programming languages
research done at Strathclyde.

\noindent Their general background and specific experience ranging from
theoretical foundations to concrete implementations mean that the
investigators are --- perhaps --- one of the few groups who could
conduct this research. They form a {\em very well balanced team} spanning
the breadth of the proposal from theory to practice and having two
implementors of programming languages - Altenkirch ($\Pi\Sigma$) and
McBride (Epigram) - will ensure robust discussions from different
perspectives which will help push the applied research forward.
Indeed, removing any of them from the project would significantly
reduce the strength of the team and the prospects for
completion. Within the team, Gambino is essential because of his
knowledge of the semantic foundations of HoTT and links with the grant
of Awodey, Altenkirch is essential for his work on the type theoretic
foundations of HoTT, Ghani is essential because of his work on logical
relations and McBride is essential for the depth of his understanding
of the implementation of functional programming languages.  As a
result, having each investigator on the proposed research is
indispensable. Of course, as we have argued, beyond their essential and unique skills,
the investigators have other overlapping skills and hence all
investigators will work closely together and be involved in all areas
of the research so that they can contribute to, and be advised of,
progress.



\vspace{0.02in}

\noindent {\bf The RAs:} The volume of work is clearly too much to be
undertaken by the investigators alone. Further, PhD students cannot be
trained in subjects like category theory, type theory, and the use and
development of program language and verification tools, and still have
enough time to conduct the proposed research. Full-time RAs are
therefore needed on this project. We had originally hoped that one RA
would suffice, but both the amount of the work and the highly distinct
skills needed between the theoretical foundations and the more
practical aspects of the project mean that we need one RA with skills
in each of these areas. In a nutshell, the volume and breadth of the
proposed research means it would be unrealistic to expect it to be
undertaken by a sole RA. The RAs will certainly need to have PhDs and,
given the diverse and specialised nature of the research, will ideally
have postdoctoral experience as well.  Such appointments are feasible
because there are a number of PhD students and RAs working in exactly
these areas, e.g., in the Coq and Agda development teams and also PhD
students who have been following the numerous summer schools that
interest in HoTT has sparked.  Although the RA will be expected to be
able to work independently, our own strengths means the RAs will be
closely managed by the investigators. They will be asked to work with
the project partners and collaborators of the investigators, to push
the project ahead; to contribute to the writing of papers and grant
proposals; and, as part of their general training, to help mentor PhD
students. A Grade 7, Point 29 salary of $\pounds$28,839 is the minimum
for newly-minted PhDs, so a Grade 7, Point 31 salary of
$\pounds$30,594 is appropriate for researchers with the qualifications
required. While it may be possible to employ new PhDs who are
appropriate for the project at Point 29, advertising a Point 31 salary
will allow us to attract researchers with more experience to the
position.


%The position will be
%required for four years to have time to fully integrate the RAs into
%the project, make progress on the proposed research, and then
%evaluate and refine the results obtained. Given the fast moving nature
%of this area of research, this length of time will allow us to absorb
%progress as it is made across the world.

\vspace{0.02in}

\noindent {\bf Duration of the Grant:} The appropriate
time required to achieve the objectives of this research is 4
years. We have proposed 8 very high quality work packages. While we
has isolated conceptually straightforward deliverables for each work
package (to ensure the project can proceed as a whole), it would be a
mistake to underestimate amount the technical work required to
formalise these concepts to the level of rigour required for a new
foundations of programming languages, robust implementations of those
foundations, and case studies detailing their use. On top of this,
we do want to tackle the more ambitious goals in our work packages
which may require conceptual
advances,  and the revision of basic working assumptions. Finally,
HoTT is a fast-moving subject and 4 years gives us time
to absorb and make use of breakthroughs likely to arise in the near
future. 
Moreover, the time required to implement and formally prove properties
within a proof system is an order of magnitude greater than for
equivalent pencil-and-paper proofs. This is the price of extra rigour.
Allocating 4 years to the
project gives us reasonable amounts of time to conclude the work in
the various packages
% --- this will be a stiff challenge, but we want to
%aim high and already have the ideas to make this feasible 
--- and also
provides sufficient time at the end of the project to serve as a
safety cushion in case any of the work packages run into unforeseen
difficulties.

%Given that four years is a reasonable amount of time to undertake the
%proposed research, one may wonder if 
\noindent Each work packages is necessary --- certainly WP1 is needed
to provide the semantic criteria against which the 
rest of our research can be judged, WP2 is needed to turn the semantic
model of HoTT from WP1 into a syntactic presentation of HoTT, WP3 is
needed since HITs are one of the exciting new features of HoTT, and
WP4 will be used to generate impact from HITs by applying them to a
significant problem in computer science, namely the theory of
effects. WP5 is needed to formally verify the correctness of our results, while WP6 and WP7 together form the programming language
we aim to develop. The value of them will be demonstrated with
concrete examples in WP8.  Thus, dropping any of the work packages
would clearly diminish the project.

\vspace{0.02in}

\noindent {\bf Hours per Week:} %We now address the expenditure
%concerning the time each of the investigators will spend on the grant
%per week. 
Each investigator will spend 20\% of their time on the project.
This means that on average they will devote one
full day per week the project. These workloads are standard for such projects, and
our experience of research management suggests any less is 
unrealistic. This allows the investigators to devote the significant
time required by the proposed research proposal while still discharging
their other responsibilities.

\vspace{0.02in}

\noindent {\bf Travel Budget:} The investigators and the RAs will
attend conferences and workshops relevant to the research:
international conferences such as LICS, ICFP, POPL, Types, TLCA, MPC,
MSFP, TFP, CSL, and ICALP. Each team member will
attend two conferences per year at an estimated average cost of
$\pounds 1000$ per trip, for a total of $\pounds 48K$. Conference
presentation and publication are the principal
means of assessing quality and disseminating
research results in computer science, and are vital to
keeping abreast of the related research of other researchers. It is
also a cost-effective mechanism for gaining high visibility for our
own research results, and for getting feedback from the community on
how to further advance our research. The cost per conference
is accurate since most conferences will be in Europe and cost
approximately $\pounds 300$ for registration, $\pounds 250$ for
travel, $\pounds 200$ for subsistence, and $\pounds 250$ for
accommodation. The cost of the occasional conference further afield
%(for example, in the US) 
will be offset by savings on occasional conferences in the UK. We
estimate that 25\% of the meetings will be within the UK and 75\% will
be international.  We have also asked for $\pounds 32K$ for four research
visits to each of our eight collaborators --- their active
participation
will provide significant added value while the costs of the 
visits are minimal when taken
over  4 years. Each visit will be for approximately one week.
and the costs are entirely standard covering travel and basic
subsistence. Finally, Dr Gambino has held a UK HoTT meeting this year
which was highly successful and Prof Ghani is scheduled to hold the
next one in Jan 2015. We will hold another workshop half way through
the grant and plan to invite each
of our collaborators as invited speakers so as to aid their formal review of the
project. This increases the quality of the review, the attractiveness of the
workshop and thus the possibilities for disseminating our results and
learning from other researchers. At the end of the grant, we will hold
a summer school to demonstrate and disseminate our results.  We have
requested \pounds 6K for the workshop and another \pounds 6K for the
summer school ... these are very small costs when taken against the 
benefits they create in terms of facilitating exchange of ideas
and generating greater impact. We also require money for travel
internally to the research team. We plan 2 group meetings per year and
request funds for 2 further trips per person per year. As these are
all UK trips, we estimate \pounds 500 per trip. Again these are
minimal costs and certainly required as face-to-face meetings can be
highly productive.


%\vspace{0.02in}
%
%\noindent {\bf Consumables:} The RA will require a PC. However, PCs
%for RAs are not provided by the University, so we request one
%here. The cost for this is $\pounds 1000$. This is an entirely
%standard request with standard pricing.

\end{document}

