\documentclass[10pt,a4paper,oneside]{letter}

\usepackage[includeheadfoot,margin=2.3cm,head=28pt,foot=20pt]{geometry}
 % \usepackage{fullpage}
 \setlength{\parindent}{0pt}

\newcommand{\entrylabel}[1]{\mbox{#1}\hfil}

\newenvironment{myitemize}
  {\begin{list}{}
    {\renewcommand{\makelabel}{\entrylabel}
    \setlength{\labelwidth}{1ex}
    \setlength{\leftmargin}{0ex}
    \setlength{\parsep}{1ex}
  }}
  {\end{list}}



\address{
{Professor Steve Awodey} \\
{Department of Philosophy} \\
{Carnegie Mellon University} \\
{Pittsburgh, USA}}

\date{June 27th, 2014}

\signature{Steve Awodey}


\begin{document}



\begin{letter}{
EPSRC \\
Polaris House, \\
North Star Avenue, \\
Swindon SN2 1ET\\
United Kingdom}

\opening{TO WHOM THIS MAY CONCERN}


I am writing in relation to the research proposal ``Homotopy Type Theory: Programming and Verification'', submitted to EPSRC by Thorsten Altenkirch, Nicola Gambino, Neil Ghani and Conor McBride.

First of all, I would like to mention that I believe this proposal to be of excellent quality and that I would be delighed to collaborate with the research team on it. 

Our collaboration would focus on the research goals of Work Package 1. The questions addressed in this work package are of fundamental importance for the future development of Homotopy Type Theory. I believe that the combination of the ability of members the research team with my own expertise would allow us to achieve its goals. Indeed, I have recently isolated a new notion of model for Homotopy Type Theory which provides a promising starting point to develop the associated model theory. 

In order to develop our collaboration, I would host 4 one-week visits by members by the research team at Carnegie Mellon University, as well as keep in regular contact by
the usual academic means.

I have already collaborated successfully with Nicola Gambino on two research papers and worked with Thorsten Altenkirch on the first book on Homotopy Type Theory, written during the special yeat at the Institute for Advanced Study in Princeton, which I co-organized in 2012/13. I know very well also the other members of the research team as well. 



As Principal Investigator of a \$7.5M grant on Homotopy Type Theory from the US Air Force Office for Scientific Research, I very much hope that research in Homotopy Type Theory will receive strong support also in the UK. 


\closing{With best regards,}

\end{letter}

\end{document}


